Data Analysis with R - Final Project
====================================
by Samit Chaudhuri

## Chemical Analysis of Red and White Wines

This project explores relationship between quality and chemical properties of
wines. Wine quality is based on subjective ratings by human experts. Chemical
properties are determined by objective measurements. Two tidy data sets of 1599
red wines and 4898 white wines, each with 11 input variables and 1 output
variables are used.

Additionally some models are presented to predict the quality of wines from
their chemical characteristics.

### Load data and necessary packages
```{r Load Red Wine Data}
redWines <- read.csv("wineQualityReds.csv")
redWines <- subset(redWines, select = -c(X))
str(redWines)

library(ggplot2)
library(GGally)
library(scales)
library(memisc)
library(gridExtra)
qualityLevel = function(quality) {
  level = as.numeric(cut(quality, breaks=c(0, 4, 5, 6, 10)))
  level
  return(level)
}
rpalette = 5
wpalette = 1
ptfont = 10           # font size of plot title
atfont = 9            # font size of axis title
ltfont = 8            # font size of legend title
```

The red wine data has one dependent variable, quality, that gives each wine a
score between 0 and 10. There are 11 independent variables, some of with may be
correlated.

```{r Quality histogram}
```

Let us first find the distribution wine quality in the sample data set. Then we
use the basic histogram to categorize quality scores into 4 levels: poor, good,
fine, and excellent. This new factor variable, quality.label, is computed by
cutting the quality number into 4 ranges (0,3], (0,5], (5,6], (6,10]. Now we can
add some color to the previous histogram.

```{r, out.width = '\\maxwidth'}
ptfont = 10           # font size of plot title
atfont = 9            # font size of axis title
ltfont = 8            # font size of legend title
rqhist1 <- ggplot(data=redWines, aes(x=quality)) +
	ggtitle("Red Wine Quality") +     # Plot title text
	xlab("Wine Quality") + ylab("Count") +  # Axis label text
    theme_bw() +                                          # Theme elements
    theme(title=element_text(size=ptfont, face="bold"),   # Title font 
         axis.title=element_text(size=atfont, face="bold")) +
    scale_x_continuous(limits=c(3,9), breaks = seq(3, 9, 1)) + 
	geom_histogram(binwidth = 1)
redWines$quality.label = cut(redWines$quality, breaks=c(0, 4, 5, 6, 10), 
    labels=c("poor", "average", "fine", "excellent")) 
rqhist2 <- ggplot(data=redWines, aes(x=quality, fill=quality.label)) +
    ggtitle("Red Wine Categories") +     # Plot title text
	xlab("Wine Quality") + ylab("Count") +  # Axis label text
    scale_fill_brewer(type = 'div', palette = rpalette, # Legend title/color
         guide=guide_legend(reverse=TRUE, title="Quality Rating")) +
    theme_bw() +                                          # Theme elements
    theme(title=element_text(size=ptfont, face="bold"),   # Title font 
        axis.title=element_text(size=atfont, face="bold"),    # Axis font 
        legend.title=element_blank(),  
        legend.position=c(.9,.7)) +                       # Legend position
    scale_x_continuous(limits=c(3,9)) + 
	geom_histogram(binwidth = 1)
grid.arrange(rqhist1, rqhist2, ncol=2)
```

So we have a decent categorization of excellent, fine, good, and poor red wines.

In the beginning, we don't really know what causes a wine to be excellent or
poor. So we want the entire data set speak for itself by plotting many variables
all at once. Hopefully the plots will draw our attention to a smaller set of
variables and relationships.

```{r, out.width = '\\maxwidth'}
ggpairs(redWines, 
		params = c(color=I('red3'), shape = I('.'), corSize=8, outlier.shape = I('.')),
		title = "Red Wines Scatter Plot Matrix")
```

Some of the dependent variables are highly correlated. For example, citric acid
is correlated with both fixed and volatile acidity. Density is correlated with
both fixed acidity and alcohol content.
 
The quality score is most correlated with alcohol content and volatile acidity,
but it does not correlate much with anything else. On the other hand, the box
plots of the quality label (excellent, fine, good, poor) show a dependency on
fix and volatile acidity; excellent wines have higher mean fixed acidity and
lower mean volatile acidity.

We will closely examine the effects of alcohol and acidity (both fixed and
volatile) on quality score and quality label with grouped density plots.

```{r, out.width = '\\maxwidth'}
ptfont = 10           # font size of plot title
atfont = 9            # font size of axis title
ltfont = 8            # font size of legend title
alDensity <- ggplot(data=redWines, aes(x=alcohol, color=quality.label)) +
	  ggtitle("Quality vs Alcohol") + # Plot title text
	  xlab("Alcohol Content") + ylab("Count") +  
      scale_color_brewer(type = 'div', palette = rpalette, # Legend title/color
          guide=guide_legend(reverse=TRUE)) +
      theme_bw() +                                          # Theme elements
      theme(title=element_text(size=ptfont, face="bold"),   # Title font 
           axis.title=element_text(size=atfont, face="bold"),   # Axis font 
           legend.title=element_blank(), 
           legend.position=c(.85,.6)) +
      geom_density() 
acidDensity <- ggplot(data=redWines) +
	  ylab("Count") +  
      scale_color_brewer(type = 'div', palette = rpalette, # Legend title/color
          guide=guide_legend(reverse=TRUE)) +
      theme_bw() +                                          # Theme elements
      theme(title=element_text(size=ptfont, face="bold"),   # Title font 
           axis.title=element_text(face="bold"),   # Axis font 
           legend.title=element_blank(), 
           legend.position=c(.85,.6))
vaDensity <- acidDensity + 
    geom_density(aes(x=volatile.acidity, color=quality.label)) + 
    ggtitle("Quality vs Volatile Acidity") +  # Plot title 
    xlab("Volatile Acidity") 
faDensity <- acidDensity + 
    geom_density(aes(x=fixed.acidity, color = quality.label)) + 
    ggtitle("Quality vs Fixed Acidity") +     # Plot title 
    xlab("Fixed Acidity") 
grid.arrange(alDensity, vaDensity, faDensity, nrow=2, ncol=2)
```

The density plots of different quality levels are shifted with respect to each
other. The alcohol density plots shift toward right with improving
quality. Better wines tend of have higher alcohol. The fixed acidity plots also
shift toward right with improving quality, while volatile acidity density plots
shift toward left with improving quality. Better wines tend of have higher fixed
acidity and lower volatile acidity.

These trends can also be observed with box plots. 

```{r, out.width = '\\maxwidth'}
ptfont = 10           # font size of plot title
atfont = 9            # font size of axis title
ltfont = 8            # font size of legend title
albox <- ggplot(data=redWines, aes(x=quality.label, y=alcohol,
	   color=quality.label)) +
	  ggtitle("Alcohol vs Quality") +     # Plot title text
	  xlab("Qality Label") + ylab("Alcohol Stats") +  # Axis label text
      scale_color_brewer(type = 'div', palette = rpalette) +# Legend title/color
      theme_bw() +                                          # Theme elements
      theme(title=element_text(size=ptfont, face="bold"),   # Title font 
           axis.title=element_text(size=atfont, face="bold"),   # Axis font 
           legend.title=element_text(size=ltfont, face="bold"), # Legend font 
           legend.position="none") +                      # Legend position
      geom_boxplot() 
abox <- ggplot(data=redWines, aes(x=quality.label, color=quality.label)) +
	       # Plot title text
	  xlab("Qality Label") +   # Axis label text
      scale_color_brewer(type = 'div', palette = rpalette)+ # Legend color
      theme_bw() +                                          # Theme elements
      theme(title=element_text(size=ptfont, face="bold"),   # Title font 
           axis.title=element_text(size=atfont, face="bold"),   # Axis font 
           legend.title=element_text(size=ltfont, face="bold"), # Legend font 
           legend.position="none")                        # Legend position
par(mfrow=c(1,2))
vabox <- abox + geom_boxplot(aes(y=volatile.acidity)) + 
	 ggtitle("Volatile Acidity vs Quality") +
     ylab("Volatile Acidity Stats")
fabox <- abox + geom_boxplot(aes(y=fixed.acidity)) +
	 ggtitle("Fixed Acidity vs Quality") +
     ylab("Fixed Acidity Stats")
grid.arrange(albox, vabox, fabox, nrow=2, ncol=2)
```

Now let's observe the joint effect of volatile acidity and alcohol content on
quality with a scatter plot. Side-by-side, we also plot the joint effect of
fixed acidity and alcohol content.

```{r}
alascatter <- ggplot(data=redWines, aes(x=alcohol, 
      group=quality.label, fill=quality.label, color=quality.label)) +
      scale_x_continuous(limits=c(8, 15), breaks = seq(8, 15, .5)) +
	  xlab("Alcohol Percentage") + # Axis label text
      scale_color_brewer(type = 'div', palette = rpalette, 
          guide=guide_legend(reverse=TRUE)) +              # Legend color
      scale_fill_brewer(guide = FALSE) +
      theme_bw() +                                         # Theme elements
      theme(title=element_text(size=ptfont, face="bold"),  # Title font 
           axis.title=element_text(size=atfont, face="bold"),   # Axis font 
           legend.title=element_blank(),                   # Legend font 
           legend.position=c(.85,.6))                      # Legend position
alvascatter <- alascatter + 
	  ggtitle("Volatile Acidity and Alcohol Content of Red Wines") + 
      ylab("Volatile Acidity") +  
      scale_y_continuous(limits=c(0.1, 1.6), breaks = seq(0.1, 1.6, .1)) +
      geom_point(aes(y=volatile.acidity, size=3))
alfascatter <- alascatter + 
	  ggtitle("Fixed Acidity and Alcohol Content of Red Wines") + 
      ylab("Fixed Acidity") +  
      scale_y_continuous(limits=c(3, 16), breaks = seq(3, 16, 1)) +
      geom_point(aes(y=fixed.acidity, size=3))
alvascatter
alfascatter
```

Red wine quality is definitely good when alcohol content is above 12 and
volatile acidity is below 0.5. Even when alcohol content is not quite as high,
one can find good wines as with low volatile acidity, or with high fixed
acidity. Similarly when volatile acidity is not quite as low, or fixed acidity
is not quite as high, one can find good wines with high alcohol content. In
other words, as long has either alcohol content is above 12, or volatile acidity
is below 8 or fixed acidity is above 8, there is a good chance that the wine has
good quality.

### Linear Regression Model of Red Wine Quality

First we start with a full linear regression model. 

```{r Regression model}
fulllm <- lm(data=redWines, quality ~ . - quality.label)
summary(fulllm)
```

Now we select only the signficant variables, and then leave out collinear
variables, citric.acid, pH, and density because they are highly correlated with
acidity.

```{r Reduced regression model}
redWines$quality = as.numeric(redWines$quality)
reducedlm <- lm(data=redWines, quality ~ . 
    - citric.acid - density - pH - quality.label)
summary(reducedlm)
```

The R-squared of the reduced model is not much different from the full
model. The R-squared did not really improve with models built with other
combinations of feature variables.

### Prediction with Classification

One possible reason that linear regression model of red-wine quality has low
accuracy is that quality has a non-linear relation with other dependent
variables. This prompts us to explore the relationship with a multi-class
classification trees. Before building a CART model, we will split the data into
a training set and a test set. CART model built on the training set will be used
to predict quality on the test set. Then we will be able to measure the accuracy
of the predictive capability of the model.

```{r, out.width = '\\maxwidth'}
library(rpart)
library(rpart.plot)
library(caret)
library(e1071)

redWines$quality = factor(redWines$quality)

# Split data into a train and test set
index <- 1:nrow(redWines)
set.seed(144)
testSamples <- sample(index, trunc(length(index)/3))
rqindex <- which(colnames(redWines)=="quality")
rqlindex <- which(colnames(redWines)=="quality.label")
redTest <- redWines[testSamples,-rqlindex]
redTrain <- redWines[-testSamples,-rqlindex]

# Create rpart model with cross validation
rtr.control = trainControl(method = "cv", number = 5)
cp.grid = expand.grid( .cp = (0:4)*0.001) 
set.seed(144)
rrpart.grid = train(data = redTrain, quality ~ ., method = "rpart", 
    trControl = rtr.control, tuneGrid = cp.grid)
rrpart.model = rrpart.grid$finalModel
prp(rrpart.model, tweak=1.5)

#Compute confusion matrix 
rrpart.cpred <- predict(rrpart.model, newdata=redTest, type = "class" )
rrpart.cmatrix <- table(pred = rrpart.cpred, true = redTest[,rqindex])
rrpart.accuracy <- sum(diag(rrpart.cmatrix))/sum(rrpart.cmatrix)
rrpart.cmatrix
rrpart.accuracy
```
Indeed, the CART model can predict quality on the test set with 58% accuracy. 

### Chemical Analysis of White Wines

We now perform EDA on white wine quality and see if they follow a similar
pattern as red wines.

```{r, out.width = '\\maxwidth'}
whiteWines <- read.csv("wineQualityWhites.csv")
whiteWines <- subset(whiteWines, select = -c(X))
str(whiteWines)

whiteWines$quality.label = cut(whiteWines$quality, breaks=c(0, 4, 5, 6, 10), 
    labels=c("poor", "average", "fine", "excellent"))
wqhist1 <- ggplot(data=whiteWines, aes(x=quality)) +
	  ggtitle("White Wine Quality") +     # Plot title text
	  xlab("Wine Quality") + ylab("Count") +  # Axis label text
      theme_bw() +                                          # Theme elements
      theme(title=element_text(size=ptfont, face="bold"),   # Title font 
           axis.title=element_text(size=atfont, face="bold")) +
      scale_x_continuous(limits=c(3,9), breaks = seq(3, 9, 1)) +  
      geom_histogram(binwidth = 1)
wqhist2 <- ggplot(data=whiteWines, aes(x=quality, fill=quality.label)) +
	      ggtitle("White Wine Categories") +     # Plot title text
	      xlab("Wine Quality") + ylab("Count") +  # Axis label text
          scale_fill_brewer(type = 'div', palette = wpalette, 
              guide=guide_legend(reverse=TRUE, title="Quality Rating")) +
          theme_bw() +                                          # Theme elements
          theme(title=element_text(size=ptfont, face="bold"),   # Title font 
             axis.title=element_text(size=atfont, face="bold"),   # Axis font 
             legend.title=element_text(size=ltfont, face="bold"), # Legend font 
             legend.position=c(.9,.7)) +                      # Legend position
		  scale_x_continuous(limits=c(3,9)) + 
		  geom_histogram(binwidth = 1)
grid.arrange(wqhist1, wqhist2, ncol=2)
``` 
Now plot all vriable pairs in the data set to identify noticable relationships.

```{r, out.width = '\\maxwidth'}
ggpairs(whiteWines, 
		params = c(color=I('green3'), shape = I('.'), corSize=8, 
            outlier.shape = I('.')),
		title = "White Wines Scatter Plot Matrix")
```

Similar to quality or red wines, quality of white wines also relate to alcohol
content and acidity. We can examine these relations with scatterplots.

```{r}
alascatter <- ggplot(data=whiteWines, aes(x=alcohol, 
      group=quality.label, fill=quality.label, color=quality.label)) +
      scale_x_continuous(limits=c(8, 15), breaks = seq(8, 15, .5)) +
	  xlab("Alcohol Percentage") +  # Axis label text
      scale_color_brewer(type = 'div', palette = wpalette,  # Legend color
          guide=guide_legend(reverse=TRUE, title="Quality Rating")) +
      scale_fill_brewer(guide = FALSE) +
      theme_bw() +                                          # Theme elements
      theme(title=element_text(size=ptfont, face="bold"),   # Title font 
           axis.title=element_text(size=atfont, face="bold"),   # Axis font 
           legend.title=element_text(size=ltfont, face="bold"), # Legend font 
           legend.position=c(.85, .6))                    # Legend position
alvascatter <- alascatter +
      ylab("Volatile Acidity") +
      scale_y_continuous(limits=c(0.08, 1.1), breaks = seq(0.08, 1.1, .1)) +
	  ggtitle("Volatile Acidity and Alcohol Content of White Wines") + 
      geom_point(aes(y=volatile.acidity), size=3)
alfascatter <- alascatter +
      ylab("Fixed Acidity") +
      scale_y_continuous(limits=c(3, 16), breaks = seq(3, 16, 1)) +
	  ggtitle("Fixed Acidity and Alcohol Content of White Wines") + 
      geom_point(aes(y=fixed.acidity), size=3)
alvascatter
alfascatter
```

We can also create a classification tree for prediction. The model can predict
quality on the test set with 52% accuracy. The CART is drawn below.

```{r, out.width = '\\maxwidth'}
library(rpart)
library(rpart.plot)
library(caret)
library(e1071)
 
whiteWines$quality = factor(whiteWines$quality)

# Split data into a train and test set
index <- 1:nrow(whiteWines)
set.seed(144)
wtestSamples <- sample(index, trunc(length(index)/3))
wqindex <- which(colnames(whiteWines)=="quality")
wqlindex <- which(colnames(whiteWines)=="quality.label")
whiteTest <- whiteWines[wtestSamples,-wqlindex]
whiteTrain <- whiteWines[-wtestSamples,-wqlindex]

# Create rpart model with cross validation
wtr.control = trainControl(method = "cv", number = 5)
wcp.grid = expand.grid( .cp = (0:4)*0.001) 
set.seed(144)
wrpart.grid = train(data = whiteTrain, quality ~ . - citric.acid - density 
    - pH , method = "rpart", trControl = wtr.control, tuneGrid = cp.grid)


wrpart.model = wrpart.grid$finalModel
prp(wrpart.model)

#Compute confusion matrix 
wrpart.cpred <- predict(wrpart.model, whiteTest, type = "class" )
wrpart.cmatrix <- table(pred = wrpart.cpred, true = whiteTest[,wqindex])
wrpart.accuracy <- sum(diag(wrpart.cmatrix))/sum(wrpart.cmatrix)
wrpart.cmatrix
wrpart.accuracy
```
```

### Interpretation of Model

More alcohol and higher quality usually go together. In addition, volatile
acidity degrades quality, probably because it induces a vinegary taste in the
wine. Residual sugar levels are somewhat important probably because sweetness is
often associated with freshness. Sulphates are important too, possibly because
of its link to fermentation.

### Final Plots and Summary

This report explored the relationship betwen chemical properties of
red and white wines, and their quality as identified by wine experts.
Several plots including histograms, density plots, box plots, and
scatter plots were used in the exploration to identify three major
featurs impacting wine quality. They are alcohol content, fixed
acidity, and volatile acidity. 

This section reproduces a few of those plots. For example, the density
plots of red wine quality over the three features are shown
below. Detailed code for generating these plots have already been
included in earlier sections.

```{r, out.width = '\\maxwidth'}
grid.arrange(alDensity, vaDensity, faDensity, nrow=2, ncol=2)
```

For white wines, the relationsip between quality, alcohol content and
volatile acidity can be seen in the following scatter plot.


```{r, out.width = '\\maxwidth'}
alvascatter
```

We also built regression and classification models to predict wine
quality from its chemical properties. The classification trees
identified two additional significant features, namely residual sugar
and sulphates. In the following we show the classification-tree for
white wines. Detailed code for generating the model have already
been included in an earlier section.

```{r, out.width = '\\maxwidth'}
prp(wrpart.model)
```
 
### Reflection

In this report, we have explored the relationship betwen chemical properties of
wines, such as fixed and volatile acidity, residual sugar, sulphates, alcohol
content, and the quality of the wines as identified by the experts. We started
with 11 different independent input variables, and showed that one can pick fine
or excellent wines by focusing on 5 or fewer primary contributors. However, our
prdictive models are not accurate enough (58% for red and 52 % for
white). Despite providing interesting insights, they need to more refined 
before providing reliable wine ratings. In addition to linear regression and
CART models used in this project, there exists other advanced techniques for
extracting features and building and training models that can be used to create
a better predictive model for wine quality.

